Effects of Reynolds Number on the Energy Conversion and Near-Wake Dynamics of a High Solidity Vertical-Axis Cross-Flow Turbine

### 2015-10-31 paper.tex
Wosnik edits

% Abstract (Less than 200 words)
\abstract{The performance and wake characteristics of cross-flow (vertical axis)
    turbines depend on many parameters: solidity, blade profile, blade pitch, number
    of blades, strut drag, operational parameters such as tip speed ratio---and
    Reynolds number. Tolerable mismatch between physical model Reynolds number and
    full-scale Reynolds number was investigated from both the perspective of
    predicting full scale performance and flow dynamics, as well as numerical model
    validation. Experiments were performed with a large laboratory-scale high
    solidity cross-flow turbine to investigate Reynolds number dependence of
    performance and wake transport characteristics, and to determine what an
    acceptable Reynolds number mismatch is for scaled physical model tests of
    cross-flow turbines. It was demonstrated that the performance of the cross-flow
    turbine becomes essentially $Re$-independent at a Reynolds number based on
    turbine diameter of $Re_D \approx 10^6$, or an approximate Reynolds number based
    on blade chord of $Re_c \approx 2 \times 10^5$. A simple model that calculates
    peak torque coefficient from static foil data and cross-flow turbine kinematics
    was shown to be a reasonable estimator for predicting Reynolds number dependence
    of an actual cross-flow turbine operating under dynamic conditions.}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical if certain nondimensional
physical parameters are matched. In the case of fluid systems, the most
important nondimensional parameter is often the Reynolds number, which
quantifies the importance of inertial forces over viscous forces on the flow
physics:
\begin{equation}
    \frac{Ul}{\nu},
    \label{eq:Re}
\end{equation}
where $U$ and $l$ are characteristic velocity and length scales, respectively,
and $\nu$ is the fluid kinematic viscosity. In general the geometric scale of a
physical model is not necessarily the same as its dynamical scale. As such,
hereafter we will use the word ``scale'' to refer to this dynamical scale or
Reynolds number.

@@

Blackwell et al. investigated the effects of Reynolds number on the performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro observed significant
Reynolds number sensitivity for a high solidity helical cross-flow turbine
\cite{Polagye2013b}, and Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$ \cite{Bravo2007}. Bachant and Wosnik \cite{Bachant2014} observed
Reynolds number independence of power coefficient at optimal tip speed ratio for
a high solidity cross-flow turbine at $Re_D \approx 10^6$ or
$Re_{c,\mathrm{avg}} \approx 2 \times 10^5$.

@@

\begin{table}
\centering
\begin{tabular}{ccc}
Tow speed (m/s) & $Re_D$ & $Re_{c,\mathrm{ave}}$ ($\lambda = 1.9$) \\
\hline
0.3 & $3.0 \times 10^5$ & $8.0 \times 10^4$ \\
0.4 & $4.0 \times 10^5$ & $1.1 \times 10^5$ \\
0.5 & $5.0 \times 10^5$ & $1.3 \times 10^5$ \\
0.6 & $6.0 \times 10^5$ & $1.6 \times 10^5$ \\
0.7 & $7.0 \times 10^5$ & $1.9 \times 10^5$ \\
0.8 & $0.8 \times 10^5$ & $2.1 \times 10^5$ \\
0.9 & $0.9 \times 10^5$ & $2.4 \times 10^5$ \\
1.0 & $1.0 \times 10^6$ & $2.7 \times 10^5$ \\
1.1 & $1.1 \times 10^6$ & $2.9 \times 10^5$ \\
1.2 & $1.2 \times 10^6$ & $3.2 \times 10^5$ \\
1.3 & $1.3 \times 10^6$ & $3.4 \times 10^5$ \\
\end{tabular}
\caption{Turbine diameter and approximate blade chord Reynolds numbers for the
tow speeds used in the experiment.}
\label{tab:Re}
\end{table}

@@

The behavior of the mean rotor drag coefficient is similar, though the changes
are less dramatic. This is likely due to cross-flow turbine geometry, where
increases in blade drag at lower $Re$ somewhat offsets the reduction in lift, to
keep total streamwise force from changing as much as the rotor torque. The
tendency for $C_D$ to continue increasing with $Re$ may also be a consequence of
increasing Froude number, which therefore increases free surface deformation and
wave drag during towing without increasing flow through the turbine.

@@

We propose a method for predicting Reynolds number dependence of a cross-flow
turbine using static airfoil characteristics and turbine blade kinematics. It is
shown that the peak torque coefficient computed this way shows similar Reynolds
number sensitivity---a better predictor than conventional quantifications of
airfoil performance, e.g., lift-to-drag ratio. The simple peak geometric torque
coefficient model predicted Reynolds number independence for Reynolds numbers
based on blade chord of $Re_c > 2 \times 10^5$. Is it also shown from this
metric that foils with higher camber are more Reynolds number dependent but have
smaller slopes in the linear regime. The Reynolds number dependence of the peak
torque coefficient calculated from static foil data and cross-flow turbine
kinematics was shown to be a reasonable indicator for predicting Reynolds number
dependence of an actual cross-flow turbine operating under dynamic conditions,
including dynamic stall.

At lower Reynolds numbers we observed lower levels of turbulence in the wake at
$x/D=1$, along with lower levels of turbulent transport with respect to mean
streamwise momentum and mean kinetic energy recovery in the streamwise
direction. Vertical advection---the largest transport mechanism
measured---showed little $Re$-dependence, whereas the negative effects of
cross-stream advection was enhanced at high $Re$. Despite being order of
magnitude smaller than other transport processes, viscous diffusion increased
rapidly with decreasing $Re$. Overall, the total wake transport measured leveled
off at essentially the same Reynolds number that the performance did.

@@

The authors would like to acknowledge funding through a National Science
Foundation CAREER award (PI Wosnik, NSF-CBET 1150797, program manager Dr. Ram
Gupta), a grant through the Leslie S. Hubbard Marine Program Endowment to
purchase acoustic flow measurement instrumentation, and a grant for laboratory
infrastructure upgrades through the US Department of Energy.

>>>>

% Abstract (Less than 200 words)
\abstract{The performance and wake characteristics of cross-flow (vertical axis)
    turbines depend on many parameters: solidity, blade profile, blade pitch, 
    number of blades, strut drag, operational parameters such as tip speed 
    ratio---and Reynolds number.  
    Experiments were performed with a large laboratory-scale high
    solidity cross-flow turbine to investigate Reynolds number dependence of
    performance and wake transport characteristics. A key question was what an
    acceptable Reynolds number mismatch is for scaled physical model tests of
    cross-flow turbines, both from the perspective of predicting full scale 
    performance and flow dynamics, as well as numerical model validation.
    It was demonstrated that the performance of the cross-flow
    turbine becomes essentially $Re$-independent at a Reynolds number based on
    turbine diameter of $Re_D \approx 10^6$, or an approximate Reynolds number based
    on blade chord of $Re_c \approx 2 \times 10^5$. A simple model that calculates
    peak torque coefficient from static foil data and cross-flow turbine kinematics
    was shown to be a reasonable estimator for predicting Reynolds number dependence
    of an actual cross-flow turbine operating under dynamic conditions.}

@@

Scaled physical models are often used in science and engineering to approximate
real-world systems. The principle of dynamic similarity allows for geometrically
scaled systems to be dynamically identical if certain nondimensional
physical parameters are matched. In the case of fluid systems, the most
important nondimensional parameter is often the Reynolds number, $Re$, which
quantifies the importance of inertial forces over viscous forces on the flow
physics:
\begin{equation}
    Re = \frac{Ul}{\nu},
    \label{eq:Re}
\end{equation}
where $U$ and $l$ are characteristic velocity and length scales, respectively,
and $\nu$ is the fluid kinematic viscosity. In general the geometric scale of a
physical model is not necessarily the same as its dynamical scale. As such,
hereafter we will use the word ``scale'' to refer to this dynamical scale or
Reynolds number.

@@

Blackwell et al. investigated the effects of Reynolds number on the performance
of a 2 m diameter Darrieus vertical-axis cross-flow wind turbine with NACA 0012
blades \cite{Blackwell1976}. By varying wind tunnel speed, the turbine was made
to operate at approximately constant blade chord Reynolds number $Re_c$ ranging
from $1 \times 10^5$ to $3 \times 10^5$. In this regime the turbine power
coefficient $C_P$ was shown to be sensitive to $Re_c$, with sensitivity
diminishing at the higher Reynolds numbers, especially for turbines of lower
solidity ($Nc/R$, where $N$ is the number of blades and $R$ is the turbine's
maximum radius). More recently, Polagye and Cavagnaro observed significant
Reynolds number sensitivity for a high solidity helical cross-flow turbine
\cite{Polagye2013b}, and Bravo et al. observed power coefficient of a
straight-bladed VAWT to become Reynolds number independent at $Re_c \approx 4
\times 10^5$ \cite{Bravo2007}. Bachant and Wosnik \cite{Bachant2014} first 
observed
Reynolds number independence of power coefficient at optimal tip speed ratio for
a high solidity cross-flow turbine at $Re_D \approx 10^6$ or
$Re_{c,\mathrm{avg}} \approx 2 \times 10^5$.

@@

\begin{table}
\centering
\begin{tabular}{ccc}
Tow speed (m/s) & $Re_D$ & $Re_{c,\mathrm{ave}}$ ($\lambda = 1.9$) \\
\hline
0.3 & $0.3 \times 10^6$ & $0.8 \times 10^5$ \\
0.4 & $0.4 \times 10^6$ & $1.1 \times 10^5$ \\
0.5 & $0.5 \times 10^6$ & $1.3 \times 10^5$ \\
0.6 & $0.6 \times 10^6$ & $1.6 \times 10^5$ \\
0.7 & $0.7 \times 10^6$ & $1.9 \times 10^5$ \\
0.8 & $0.8 \times 10^6$ & $2.1 \times 10^5$ \\
0.9 & $0.9 \times 10^6$ & $2.4 \times 10^5$ \\
1.0 & $1.0 \times 10^6$ & $2.7 \times 10^5$ \\
1.1 & $1.1 \times 10^6$ & $2.9 \times 10^5$ \\
1.2 & $1.2 \times 10^6$ & $3.2 \times 10^5$ \\
1.3 & $1.3 \times 10^6$ & $3.4 \times 10^5$ \\
\end{tabular}
\caption{Turbine diameter and approximate blade chord Reynolds numbers for the
tow speeds used in the experiment.}
\label{tab:Re}
\end{table}

@@

The behavior of the mean rotor drag coefficient is similar, though the changes
are less dramatic. This is likely due to cross-flow turbine geometry, where
increases in blade drag at lower $Re$ somewhat offsets the reduction in lift, to
keep total streamwise force from changing as much as the rotor torque. The
tendency for $C_D$ to continue increasing with $Re$ may also include the 
effects of an 
increasing Froude number, which increases free surface deformation and
wave drag during towing without increasing flow through the turbine.

@@

We propose a method for predicting Reynolds number dependence of a cross-flow
turbine using static airfoil characteristics and turbine blade kinematics. It is
shown that the peak torque coefficient computed this way shows similar Reynolds
number sensitivity---a better predictor than conventional quantifications of
airfoil performance, e.g., lift-to-drag ratio. The simple peak geometric torque
coefficient model predicted Reynolds number independence for Reynolds numbers
based on blade chord of $Re_c > 2 \times 10^5$. Is it also shown from this
metric that foils with higher camber are more Reynolds number dependent at 
lower Reynolds number but have
smaller slopes in the linear regime. The Reynolds number dependence of the peak
torque coefficient calculated from static foil data and cross-flow turbine
kinematics was shown to be a reasonable indicator for predicting Reynolds number
dependence of an actual cross-flow turbine operating under dynamic conditions,
including dynamic stall.

At lower Reynolds numbers we observed lower levels of turbulence in the wake at
$x/D=1$, along with lower levels of turbulent transport with respect to mean
streamwise momentum and mean kinetic energy recovery in the streamwise
direction. Vertical advection---the largest transport mechanism
measured---showed little $Re$-dependence, whereas the negative effects of
cross-stream advection was enhanced at high $Re$. Despite being order of
magnitude smaller than other transport processes, viscous diffusion increased
rapidly with decreasing $Re$, and is expected to become significant to the wake 
dynamics around $Re_D \sim 10^4$.
Overall, the total wake transport measured leveled
off at essentially the same Reynolds number that the performance did.

@@

The authors would like to acknowledge funding through a National Science
Foundation CAREER award (PI Wosnik, NSF-CBET 1150797, program manager Dr. 
Gregory L. Rorrer), a grant through the Leslie S. Hubbard Marine Program 
Endowment to
purchase acoustic flow measurement instrumentation, and a grant for laboratory
infrastructure upgrades through the US Department of Energy.
